---
output:
  md_document:
    variant: markdown_github
    toc: FALSE
---

##American political data & R: some open source resources & methods

An R-based guide to accessing, exploring & visualizing US political data utilizing a collection of open government resources, including presidential election returns (2008-2016 by congressional district), lawmaker biographies & political ideologies, and congressional district demographics.

Data used in this guide have been collated from [The Daily Kos](https://www.dailykos.com/stories/2018/2/21/1742660/-The-ultimate-Daily-Kos-Elections-guide-to-all-of-our-data-sets), [CivilServiceUSA](https://github.com/CivilServiceUSA), and the R packages [tidycensus](https://walkerke.github.io/tidycensus/) & [Rvoteview](https://github.com/voteview/Rvoteview).


- [1 Lawmaker details](#1-Lawmaker-details)
- [2 Political Ideologies](#2-political-ideologies-and-congressional-composition)
- [3 Political geometries](#3-political-geometries)
- [4 Federal election results](#4-Federal-election-results)
- [5 Census data and congressional districts](#5-Census-data-and-congressional-districts)
- [6 Equal-area political geometries](#6-Equal-area-political-geometries)
- [7 Summary](#7-Summary)


Hopefully a useful **open source & transparent framework** for investigating past & future election results and congresses using R.  All work presented here can be reproduced in its entirety. A developing resource. Open government data.


```{r message=FALSE, warning=FALSE}
library(tidyverse)
```


---

### 1 Lawmaker details

> [CivilServiceUSA](https://github.com/CivilServiceUSA) provides a wonderful collection of details about each lawmaker in the 116th Congress, including age, race, religion, biographical details, and social media info.  A full roll call of information available for each lawmaker is available [here](https://github.com/CivilServiceUSA/us-house#data-set). Here, we consider some different perspectives on the composition of the 116th House utilizing these data.


```{r message=FALSE, warning=FALSE}
git <- 'https://raw.githubusercontent.com/CivilServiceUSA/'
csusa_house_dets <- jsonlite::fromJSON(url(paste0(git, 'us-senate/master/us-senate/data/us-senate.json')))
csusa_house_dets <- jsonlite::fromJSON(url(paste0(git, 'us-house/master/us-house/data/us-house.json')))
```



> A small sample of the data made available by CivilServiceUSA:

```{r}
set.seed(9997)
csusa_house_dets %>%
  sample_n(5) %>%
  select(name, state_code, district, party, gender, ethnicity, twitter_handle) %>%
  knitr::kable(row.names = FALSE)
```



#### 1.1 Age & generational demographics of the 116th House

> An overview of the new 116th House in terms of age and generational distributions.    

```{r fig.height=4.5, message=FALSE, warning=FALSE}
csusa_house_dets %>%
  mutate (years = 
            lubridate::year(as.Date(Sys.Date())) -
            lubridate::year(as.Date(date_of_birth))) %>%
  ggplot (aes(years)) +
  geom_histogram(bins=20, fill = 'steelblue', alpha = .85) +
  labs(title = '116th House composition by age',
       caption = 'Data source: CivilServiceUSA')
```


> [Pew Research](http://www.pewresearch.org/fact-tank/2018/04/11/millennials-largest-generation-us-labor-force/ft_15-05-11_millennialsdefined/) uses the following set of birth-year ranges to delineate generations.  

* Millenials: 1981-1997
* Generation X: 1965 -1980
* Baby Boomers: 1946-1964
* Silent: 1928-1945
* Greatest: < 1928


> For good measure, we provide a more detailed classification of Boomers --- lumping folks born post-WWII with those born in the ~Sixties is a problem for me. So, (a) Boomers-proper 1946-1954 & (b) [Generation Jones](https://en.wikipedia.org/wiki/Generation_Jones) 1955-1964. 


```{r fig.height=3.5}
gens116 <- csusa_house_dets %>%
  mutate (yob = as.numeric(gsub('-.*$', '', date_of_birth))) %>%
  mutate (gen = case_when (yob < 1998 & yob > 1980 ~ '4- Millenial',
                           yob < 1981 & yob > 1964 ~ '3- Gen X',
                           yob < 1965 & yob > 1954 ~ '2b - Gen Jones',
                           yob < 1955 & yob > 1945 ~ '2a - Boomer-proper',
                           yob < 1946 & yob > 1927 ~ '1 - Silent'))
gens116 %>%
  filter(!is.na(gen)) %>%
  group_by(gen,party) %>%
  summarize(n=n()) %>%
  group_by(party) %>%
  mutate(rank = row_number())%>%
  ggplot(aes(x=reorder(gen, -rank), 
             y=n, 
             fill=gen)) + 
  geom_col(show.legend = FALSE, alpha = 0.85)+
  geom_text(aes(label = n), size = 3)+
  ggthemes::scale_fill_stata() +
  xlab(NULL) + ylab(NULL) +
  facet_wrap(~party) +
  coord_flip() +
  labs(title = '116th US House composition by generation',
       caption = 'Data source: CivilServiceUSA')
```




#### 1.2 Ethnicity & gender in the 116th House

> A quick look at gender & ehtnicity diversity by party affiliation in the new 116th House.

```{r fig.height=6}
csusa_house_dets %>%
  group_by(party, ethnicity, gender) %>%
  summarize(n = n()) %>%
  #na.omit() %>%
    ggplot(aes(area = n,
               fill = ethnicity,
               label = gender,
               subgroup = toupper(ethnicity))) +
      treemapify::geom_treemap(alpha=.85) +
      treemapify::geom_treemap_subgroup_border() +
      treemapify::geom_treemap_text(colour = "white", 
                        place = "topleft", 
                        reflow = T,
                        size = 11)+
      treemapify::geom_treemap_subgroup_text(place = "bottom", 
                                 grow = F, 
                                 alpha = 0.65, 
                                 colour ="black",  
                                 size = 13)+   
      ggthemes::scale_fill_stata()+
      facet_wrap(~party)+
      theme(legend.position = "bottom",
            legend.title=element_blank()) +
      labs(title = '116th House composition by party, ethnicity & gender',
           caption = 'Data source: CivilServiceUSA')
```


---

### 2 Political ideologies and congressional composition

> The [VoteView](https://voteview.com/) project provides roll call-based political ideology scores for all lawmakers in the history of the US Congress.  Data can be used to investigate congressional composition by party affiliation over time, the aggregate political ideologies of both houses over time, and the ideologies of individual lawmakers. And any number of other roll call-based analyses. The R package `Rvoteview` provides access to these data.


```{r message=FALSE, warning=FALSE}
rvoteview_house_50 <- lapply(c(66:116), function (x)
                    Rvoteview::member_search (
                      chamber = 'House', 
                      congress = x)) %>% 
  bind_rows() %>%
  mutate(name = paste0 (gsub(', .*$', '', bioname), ' ', cqlabel))
```




#### 2.1 Congressional composition by political affiliation


> A summary of House compositions for the last fifty congresses, ie, last 100 years. Until fairly recently, a Democratic stronghold.


```{r fig.height=5.5, message=FALSE, warning=FALSE}
rvoteview_house_50 %>%
  filter(party_name %in% c('Democratic Party', 'Republican Party') &
      congress > 66) %>%
  group_by(congress, party_name) %>%
  summarize(n = n()) %>%
  mutate(n = n/sum(n)) %>%
  ggplot(aes(x=congress, y=n, fill = party_name)) +
  geom_area(alpha = 0.85, color = 'gray') +
  ggthemes::scale_fill_stata()+
  geom_hline(yintercept = 0.5, color = 'white', linetype = 2) +
  annotate("text", x = 73, y = .9, label = "Democrats", 
           color = 'white',
           size = 4) +
  annotate("text", x = 73, y = .1, label = "Republicans", 
           color = 'white',
           size = 4) +
  theme(legend.position = "none")+
  labs(title = "House Composition over the last 50 congresses (67 to 116)",
       caption = 'Data source: VoteView')
```


#### 2.2 Lawmaker political ideologies

> The [NOMINATE scaling procedure](https://voteview.com/about) is used to calculate the political ideology of lawmakers based on voting behavior.  Ideologies are scored along two dimensions.  The first captures ideological variation based in the standard liberal-conservative dvide. The second captures variation based in social conservatism that crosscuts political affiliation.  

```{r}
extremes <- rvoteview_house_50 %>%
  filter(congress == 111) %>%
  select(name, party_name, nominate.dim1, nominate.dim2) %>%
  gather (key = dim, value = estimate,
          nominate.dim1:nominate.dim2) %>%
  group_by(party_name,dim) %>%
  filter(estimate == max((estimate)) | 
           estimate == min((estimate)))
```


> The 111th House took office in 2009; members were elected in a Presidential election year with Obama & McCain on the top of the ticket.  Below we plot members of the 111th House in two-dimensional political space.  Ideoligically extreme members are labeled.

```{r fig.height=5, message=FALSE, warning=FALSE}
rvoteview_house_50 %>%
  filter (congress == 111) %>%
  ggplot(aes(x=nominate.dim1, 
             y=nominate.dim2, 
             label = name
             )) +
          annotate("path",
               x=cos(seq(0,2*pi,length.out=300)),
               y=sin(seq(0,2*pi,length.out=300)),
               color='gray',
               size = .25) +
  geom_point(aes(color = as.factor(party_code)), 
             size= 2, 
             shape= 17) +
  ggrepel::geom_text_repel(
    data  = filter(rvoteview_house_50, 
                   congress==111,
                   name %in% extremes$name),
    nudge_y =  -0.025,
    direction = "y",
    hjust = 0, size = 2.5) +
  ggthemes::scale_color_stata() +
  theme(legend.position = 'none') +
  labs(title="DW-Nominate ideology scores for the 111th US House",
       caption = 'Data source: VoteView')
```




#### 2.3 Political ideologies historically: a party-based overview

> A slightly modified version of [this](https://voteview.com/parties/all) VoteView visualization illustrating the growing ideological divide between major political parties in the US. The gray line reflects overall median ideology scores in the House, and roughly approximates party control of the House historically.

```{r fig.height=5.5}
rvoteview_house_50 %>%
  filter(!is.na(nominate.dim1) & 
           party_name %in% c('Democratic Party','Republican Party')&
           congress > 66) %>%
  group_by(congress) %>%
  mutate(xmed = median(nominate.dim1)) %>%
  group_by(congress, party_name, xmed) %>%
  summarize(med = median(nominate.dim1)) %>%
  ungroup() %>%
  mutate(year = 1917 + 2*rep(c(1:50), each = 2)) %>%
  ggplot() +
  geom_line(aes(x = year, y= med, color = party_name), size = 1) +
  ggthemes::scale_color_stata()+
  geom_line(aes(x = year, y= xmed), color = 'darkgray',  size = 1.25) +
  theme(legend.position = 'bottom') +
  labs(title="Median ideologies for major parties: Houses 66 to 115",
       caption = 'Data source: VoteView') 
```




#### 2.4 NOKKEN & POOLE scores

> An alternative set of ideology scores are available as a csv from VoteView's website.  DW-Nominate scores are constant over a lawmakers entire political career.  Nokken & Poole scores, in contrast, are congress-specific scores.


```{r message=FALSE, warning=FALSE, eval=FALSE}
voteview_nokken_poole <- read.csv(url("https://voteview.com/static/data/out/members/HSall_members.csv"),
  stringsAsFactors = FALSE) 
```


---


### 3 Political geometries

> The R package `tigris` provides access to all US Census-based geometries, including US congressional districts and state legislative districts.

```{r message=FALSE, warning=FALSE}
nonx <- c('78', '69', '66', '72', '60', '15', '02')

library(tigris); options(tigris_use_cache = TRUE, tigris_class = "sf")
us_house_districts <- tigris::congressional_districts(cb = TRUE) %>%
  select(GEOID,STATEFP, CD115FP) %>%
  
  left_join(tigris::states(cb = TRUE) %>% 
              data.frame() %>%
              select(STATEFP, STUSPS)) 

laea <- sf::st_crs("+proj=laea +lat_0=30 +lon_0=-95") # Lambert equal area
us_house_districts <- sf::st_transform(us_house_districts, laea)
```

  
---

### 4 Federal election results

> The Daily Kos makes available [a host of data sets](https://www.dailykos.com/stories/2018/2/21/1742660/-The-ultimate-Daily-Kos-Elections-guide-to-all-of-our-data-sets), including federal election returns for presidential & congressional races.  2016 returns are the most recent data sets available. Here we focus on presidential results for the last three elections.  Seemingly one of the more comprehensive collections of open source federal election returns.



#### 4.1 Restructuring election data

```{r message=FALSE, warning=FALSE}
url <- 'https://docs.google.com/spreadsheets/d/1oRl7vxEJUUDWJCyrjo62cELJD2ONIVl-D9TSUKiK9jk/edit#gid=1178631925'
house <- gsheet::gsheet2tbl(url) 
```



> **A simple cleaning procedure** that should (more or less) scale to other data sources at the Daily Kos. 

```{r message=FALSE, warning=FALSE}
fix <- as.data.frame(cbind(colnames(house), as.character(house[1,])), 
  string_as_factor = FALSE) %>%
  mutate(V1 = gsub('^X', NA, V1)) %>%
  fill(V1) %>%
  mutate(nw_cols = ifelse(is.na(V2), V1, paste0(V1, '_', V2)),
         nw_cols = gsub(' ', '_', nw_cols))

colnames(house) <- fix$nw_cols
house <- house %>% slice(3:nrow(.))
keeps <- house[,!grepl('Pronun|ACS|Census|Survey', colnames(house))]
```



> Our new data structure, filtered to Presidential returns.

```{r message=FALSE, warning=FALSE}
dailykos_pres_elections <- keeps [,c('District', 'Code', grep('President_[A-z]', colnames(house), value=T))] %>%
  gather (key = election, value = percent, `2016_President_Clinton`:`2008_President_McCain`) %>%
  mutate(election = gsub('President_', '', election),
         percent = as.numeric(percent)) %>%
  separate(Code, c('STUSPS', 'CD115FP')) %>%
  separate(election, c('year', 'candidate'))%>%
  mutate(CD115FP = ifelse(CD115FP == 'AL', '00', CD115FP)) %>%
  left_join(data.frame(us_house_districts) %>% select (-geometry))
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
dailykos_pres_elections %>%
  select(District, GEOID, year, candidate, percent) %>%
  head() %>%
  knitr::kable()
```




#### 4.2 Presidential election results by district - 2016

> **Trump vote margin** by congressional district = Trump vote share (%) - Clinton vote share (%)

```{r fig.height=6, message=FALSE, warning=FALSE}
us_house_districts %>%
  filter(!gsub('..$' ,'', GEOID) %in% nonx) %>%
  left_join(dailykos_pres_elections %>% 
              filter(year == '2016') %>%
              spread(candidate, percent) %>%
              mutate(Margin = Trump-Clinton)) %>%
  ggplot() + 
  geom_sf(aes(fill = Margin)) +
  scale_fill_distiller(palette = "RdBu",direction=-1)+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        legend.position = 'bottom') +
  labs(title = "Trump vote margins by congressional district",
       caption = 'Data source: Daily Kos')
```



#### 4.3 Rural & urban voting

> Using the area of congressional districts (in log square meters) as a proxy for the often cited **urban-rural divide** in American votership, the plot below illustrates the relationship between Trump margins and ~degree of urbanicity.  The pattern below helps account for the very red map above despite Clinton's success in the popular vote.

```{r fig.height=5, message=FALSE, warning=FALSE}
us_house_districts %>%
  left_join(dailykos_pres_elections %>% 
              filter(year == '2016') %>%
              spread(candidate, percent) %>%
              mutate(Margin = Trump-Clinton)) %>%
  mutate(area = as.numeric(gsub(' m^2]', '', sf::st_area(.))),
         party = ifelse(Margin > 0, 'Trump', 'Clinton')) %>%
  ggplot(aes(y=Margin, x=log(area), color = party)) +
  geom_point(alpha = .65) +
  ggthemes::scale_color_stata()+
  geom_smooth(method="lm", se=T, color = 'black', linetype = 3)+
  geom_hline(yintercept = 0, color = 'darkgrey') +
  theme(legend.position = "none")+
  labs(title = "Trump vote margins vs. log(area) of congressional district",
       caption = 'Data source: Daily Kos')
```


---

### 5 Census data and congressional districts

> The US Census/American Community Survey (ACS) make counts/estimates available by congressional district.  The R package `tidycensus` provides access to census APIs.  Here we gather educational attainment data.  

```{r message=FALSE, warning=FALSE, include=FALSE}
var_list <- tidycensus::load_variables(2017, "acs5")
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
tidycensus::census_api_key("b508704c99f3ae9bc5b5e7c41e3dd77e59d52722")
```



#### 5.1 Educational attainment by race/ethnicity

```{r message=FALSE, warning=FALSE}
code <- c('A', 'B', 'C', 'D', 'E',
          'F', 'G', 'H', 'I')
          
          
race <- c('WHITE ALONE', 'BLACK OR AFRICAN AMERICAN ALONE',
          'AMERICAN INDIAN OR ALASKAN NATIVE ALONE',
          'ASIAN ALONE', 
          'NATIVE HAWAIIAN AND OTHER PACIFIC ISLANDER ALONE', 
          'SOME OTHER RACE ALONE', 'TWO OR MORE RACES',
          'WHITE ALONE, NOT HISPANIC OR LATINO',
          'HISPANC OR LATINO')

race_table <- as.data.frame(cbind(code,race),
                            stringsAsFactors=FALSE)
```



> Table C15002: *Sex by educational attainment for the population 25 years and over*.  Data disaggregated by race & ethncity can be accessed via suffixes A-I. The table above crosses suffix to subgroup classification.

```{r message=FALSE, warning=FALSE}
search_vars <- var_list[grepl('C1500', var_list$name),]

tidycens_data <- tidycensus::get_acs(geography = 'congressional district',
                            variables = search_vars$name,
                            summary_var = 'B15002_001',
                            year = 2017,
                            survey = 'acs5') %>%
  left_join(search_vars %>% rename(variable = name)) %>%
  filter(!grepl('Total$|Female$|Male$', label)) %>%
  
  mutate(gender = ifelse(grepl('Male', label), 'Male', 'Female'),
         label = gsub('^Estimate.*!!', '', label),
         code = gsub('(C[0-9]+)([A-Z])(_[0-9]+.$)', 
                     '\\2', 
                     variable)) %>%
  left_join (race_table) %>%
  select(GEOID, label, gender, race, estimate:summary_moe)
```



#### 5.2 The White working class

> In the Trump era, educational divides in voting behavior have received a great deal of attention.  In particular, pundits & news outlets have focused on a segment of the population they have dubbed the "White working class."  

> **White working class** formalized in US Census terms: Population 25 years & older who identify as White & non-Hispanic without a Bachelor's degree.


```{r message=FALSE, warning=FALSE}
us_house_districts %>% 
  filter(!gsub('..$' ,'', GEOID) %in% nonx) %>%
  left_join(tidycens_data %>% 
              filter(label != 'Bachelor\'s degree or higher' &
                     #gender == 'Male' & 
                     race == 'WHITE ALONE, NOT HISPANIC OR LATINO')%>%
              group_by(GEOID) %>%
              summarize(estimate = sum(estimate), 
                        summary_est = mean(summary_est))) %>%
  mutate(per = estimate / summary_est) %>%
  ggplot() + 
  geom_sf(aes(fill = per)) + 
  scale_fill_distiller(palette = "BrBG", direction=1)+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        legend.position = 'bottom') +
  labs(title = "White working class (%) by congressional district",
       caption = 'Data source: ACS 5-Year estimates, 2013-17, Table C15002')
```




#### 5.3 Race & educational attainment profiles by district

> Profiling congressional districts in terms of race/ethncitiy & educational attainment. Categories include: 


* White with college degree, 
* White without college degree, 
* Non-White with college degree, and 
* Non-White without college degree.  


> The **Non-White** population is defined more specifically as non-White (race) and/or Hispanic (ethnicity).


```{r message=FALSE, warning=FALSE}
tree <- tidycens_data %>%
  left_join(data.frame(us_house_districts) %>% select(GEOID, STUSPS, CD115FP)) %>%
  mutate (race = gsub(', | ', '_', race)) %>%
  select(-moe:-summary_moe) %>%
  spread(race, estimate) %>%
  mutate(WHITE_ALONE_HISPANIC = WHITE_ALONE - WHITE_ALONE_NOT_HISPANIC_OR_LATINO) %>%
  gather(key =race, value = estimate, AMERICAN_INDIAN_OR_ALASKAN_NATIVE_ALONE:WHITE_ALONE_HISPANIC) %>%
  filter(race != 'HISPANIC OR LATINO') %>%
  mutate(race_cat = ifelse(race == 'WHITE_ALONE_NOT_HISPANIC_OR_LATINO', 'White', 'Non-White'),
    ed_cat = ifelse(label == 'Bachelor\'s degree or higher', 'College', 'Non-College'))%>%
  group_by(GEOID, STUSPS, CD115FP, race_cat, ed_cat) %>%
  summarize(estimate = sum(estimate)) %>%
  group_by(GEOID) %>%
  mutate(per = estimate/sum(estimate)) %>%
  ungroup()
```


> **Race by educational attainment profiles** for a random sample of congressional districts:

```{r fig.height=5.5, message=FALSE, warning=FALSE}
set.seed(99)
samp_n <- sample(unique(tree$GEOID), 12)

tree %>%
  filter(GEOID %in% samp_n) %>%
    ggplot(aes(area = per,
               fill = paste0(race_cat, ' ', ed_cat),
               label = paste0(race_cat, ' ', ed_cat),
               subgroup = paste0(race_cat, ' ', ed_cat)))+
      treemapify::geom_treemap(alpha=.7)+
      treemapify::geom_treemap_subgroup_border() +

      treemapify::geom_treemap_text(colour = "black", 
                        place = "topleft", 
                        reflow = T,
                        size = 8)+
      #ggthemes::scale_fill_stata()+ 
      scale_fill_brewer(palette = 'Paired') +
      facet_wrap(~paste0(STUSPS, '-', CD115FP)) +
      theme(legend.position = "none",
            legend.title=element_blank()) + 
      labs(title = "Race by educational attainment profiles",
           subtitle = 'A random sample of congressional districts',
           caption = 'Source: ACS 5-Year estimates, 2013-17, Table C15002')
```



#### 5.4 White working class & Republican vote shares historically

> A comparison of the relationship between the share of working class Whites and share of Republican candidate support by district for the last three presidential elections.

```{r fig.height=6, message=FALSE, warning=FALSE}
by_pres <- dailykos_pres_elections %>%
  filter(candidate %in% c('McCain', 'Romney', 'Trump')) %>% 
  left_join(tree %>% filter(race_cat == 'White' & 
                              ed_cat == 'Non-College')) %>%
  rename(Per_White_Working = per, Per_Rep_Share = percent)
  
ggplot(data = by_pres, aes(x=Per_White_Working, 
                           y=Per_Rep_Share, 
                           fill = candidate, 
                           color = candidate, 
                           linetype = candidate)) +
  geom_point(alpha = .75, size = 1) +
  ggthemes::scale_fill_stata()+
  ggthemes::scale_color_stata()+
  geom_smooth(method="lm", se=T, color = 'black', size = .5)+
  theme(legend.position = "bottom")+
  labs(title = "% White working class vs. % Republican vote share",
       subtitle = 'By Presidential election',
       caption = 'Data source: Daily Kos & ACS')
```


> **As the table below attests**, this relationship has grown stronger in the Trump era.

```{r}
by_pres %>% 
  group_by(candidate) %>%
  summarize(cor_coefficient = cor(Per_White_Working, Per_Rep_Share)) %>%
  knitr::kable()
```


---

### 6 Equal-area political geometries

> The Daily Kos makes available a set of alternative geometries that spatially represent congressional districts and states as equal-area polygons.  As some of the maps presented above attest, America's larger states and congressional districts tend to overwhelm a standard map, and ultimately serve to under-represent/hide smaller, more populous urban areas. 

> **Links** to these shapefiles are detailed below:


```{r}
base <- 'https://drive.google.com/uc?authuser=0&id='
#Hex map
dailyvos_hex_cd <- paste0(base, '1E_P0r1Uv438fZsvKsvidIR02Nb5Ju9zf&export=download/HexCDv12.zip')
dailyvos_hex_st <- paste0(base, '0B2X3Bx1aCHsJVWxYZGtxMGhrMEE&export=download/HexSTv11.zip')
#Tile map
dailyvos_tile_outer <- paste0(base, '0B2X3Bx1aCHsJdGF4ZWRTQmVyV2s&export=download/TileOutv10.zip')
dailyvos_tile_inner <- paste0(base, '0B2X3Bx1aCHsJR1c0SzNyWlAtZjA&export=download/TileInv10.zip')
```



> A simple function for shapefile extraction:

```{r message=FALSE, warning=FALSE}
get_url_shape <- function (url) {
  temp <- tempdir()
  zip_name <- paste0(temp, '\\', basename(url))
  download.file(url, zip_name, 
                quiet = TRUE)
  unzip(zip_name, exdir = temp)
  x <- sf::st_read(dsn = gsub('\\.zip', '', zip_name), 
                   layer = gsub('\\.zip','', basename(url)),
                   quiet = TRUE) 
  unlink(temp) 
  x}
```



#### 6.1 Tile map of US states: Senate composition historically

```{r}
dailykos_tile <- lapply (c(dailyvos_tile_inner,
                           dailyvos_tile_outer),
                         get_url_shape)
names(dailykos_tile) <- c('inner', 'outer')
```


```{r message=FALSE, warning=FALSE, include=FALSE}
rvoteview_senate_50 <- lapply(c(66:115), function (x)
                    Rvoteview::member_search (
                      chamber = 'Senate', 
                      congress = x)) %>% 
  bind_rows() 
```


> An overview of **Senate representation** by state and party affiliation utilizing an equal-area tile map of US states. 

```{r}
sens <- rvoteview_senate_50 %>%
  mutate(party_name = factor(party_name, levels =c('Democratic Party', 
                                         'Republican Party', 
                                         'Independent'))) %>%
  filter(congress %in% c(85, 91, 97, 103, 109, 115)) %>%
  arrange (state_abbrev, party_name) %>%
  group_by(congress, state_abbrev) %>%
  mutate(layer = row_number())%>%
  rename(State = state_abbrev) %>%
  select(congress, State, party_name, layer)
```



> The **tile map** below illustrates the evolution of US Senate composition by state and party affiliation over the last thirty congresses. Indeed a shifting landscape.

```{r fig.height=6, message=FALSE, warning=FALSE}
dailykos_tile$outer %>% 
  left_join(sens %>% filter (layer == 2)) %>%
  ggplot() + 
  geom_sf(aes(fill = party_name),
          color = 'black', 
          alpha = .85) + 
  geom_sf(data = dailykos_tile$inner %>%
            left_join(sens %>% filter (layer == 1)), 
          aes(fill = party_name)) +
  ggsflabel::geom_sf_text(data = dailykos_tile$inner,
                          aes(label = State), 
                          size = 1.75,
                          color = 'white') +
  ggthemes::scale_fill_stata()+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        legend.title=element_blank(),
        legend.position = 'bottom') +
  facet_wrap(~congress) +
  labs(title = "US Senate Composition by State & Party in 6 congressional snapshots",
       caption = 'Data sources: Daily Kos & VoteView')
```


> Counts of states with **split Senate delegations** by congress/year.  Split delegation = Senators from the same state with different party affiliations.   

```{r fig.height=4}
rvoteview_senate_50 %>%
  filter(congress > 84) %>%
  group_by(congress, state) %>%
  summarize(splits = length(unique(party_code))) %>%
  filter(splits == 2) %>%
  group_by(congress) %>%
  summarize(n=n())%>%
  mutate(year = 1955 + 2*rep(c(1:31))) %>%
  ggplot() +
  geom_line(aes(x = year, y= n), size = 1.5, color= 'steelblue')+
  ylim(10,30)+
  labs(title = "Split delegations in the US Senate: Congresses 85 to 115",
       caption = 'Data sources: VoteView')
```




#### 6.2 Hexmap of Congressional districs: Presidential voting groups

> A equal-area/hexmap perspective on the voting patterns of congressional districts for the last three presidential elections. 

```{r message=FALSE, warning=FALSE}
dailykos_shapes <- lapply (c(dailyvos_hex_cd, dailyvos_hex_st), 
                           get_url_shape)
names(dailykos_shapes) <- c('cds', 'states')
#State hex shapefile is slightly broken.
dailykos_shapes$states <- lwgeom::st_make_valid(dailykos_shapes$states)
```



> Here we treat presidential winners in each district for election years 2008, 2012, 2016 as a single "series".  A total of 2 x 2 x 2 = 8 series, or voting groups, are theoretically possible.    


```{r}
dailykos_pres_flips <- dailykos_pres_elections %>%
  group_by(District, year) %>%
  filter(percent == max(percent))%>%
  
  mutate(dups = n()) %>%
  filter(dups != 2) %>% #Kill ties --> n = 3
  
  select(-percent, -dups) %>% #
  arrange(District) %>%
  spread(year, candidate) 

#Some hand edits to address percentage "ties" 
dailykos_pres_flips$`2012`[dailykos_pres_flips$District == 'Florida 7th'] <- 'Obama' 
dailykos_pres_flips$`2008`[dailykos_pres_flips$District == 'Ohio 10th'] <- 'Obama' 
dailykos_pres_flips$`2008`[dailykos_pres_flips$District == 'New York 22nd'] <- 'McCain'

dailykos_pres_flips <- dailykos_pres_flips %>%
  mutate(flips = paste0(`2008`, '-',`2012`, '-', `2016`)) %>%
  group_by(flips) %>%
  mutate(sum = n()) %>%
  ungroup()
```


> A summary of congressional district counts by presidential voting group for the 2008, 2012 & 2016 presidential elections. Only the series **McCain < Obama < Clinton** is unattested.

```{r fig.height=3.5, message=FALSE, warning=FALSE}
dailykos_pres_flips %>%
  group_by(flips) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  ggplot(aes(x=reorder(flips, n), 
             y=n, 
             fill=reorder(flips, -n))) + 
  geom_col(show.legend = FALSE, alpha = 0.85)+
  geom_text(aes(label = n), size = 3)+
  ggthemes::scale_fill_stata() +
  xlab(NULL) + ylab(NULL) +
  coord_flip() +
  labs(title = 'Presidential voting groups - 2008-16',
       caption = 'Data source: Daily Kos')
```



> An **equal-area map of congressional districts** illustrating voting groups for the 2008, 20012 & 2016 presidential elections.

```{r fig.height=6, message=FALSE, warning=FALSE}
dailykos_shapes$cds %>%
  inner_join(dailykos_pres_flips)%>%
  ggplot() + 
  geom_sf(aes(fill = reorder(flips, -sum)),
          color = 'gray', 
          alpha = .85) + 
  geom_sf(data=dailykos_shapes$states, 
          fill = NA, 
          show.legend = F, 
          color="black", 
          lwd=.7) +
  ggsflabel::geom_sf_text(data = dailykos_shapes$states,
                          aes(label = STATE), 
                          size = 2.75,
                          color='white',
                          face='bold') +
  ggthemes::scale_fill_stata()+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        legend.title=element_blank(),
        legend.position = 'bottom') +
  labs(title = "Presidential voting groups by district - 2008, 2012 & 2016",
       caption = 'Data source: Daily Kos')
```




#### 6.3 Presidential voting groups in 2-steps: A Sankey perspective


> Presidential voting groups as a two-step series. A more detailed perspective.

```{r}
dailykos_pres_flips %>%
  mutate(f08_12 = paste0(`2008`,'_', `2012`),
         f12_16 = paste0(`2012`,'_', `2016`))%>%
  select(District, f08_12, f12_16) %>%
  gather(elect, flip, -District) %>%
  group_by(elect, flip) %>%
  summarize(value=n()) %>%
  separate(flip, c('source', 'target'), sep = '_') %>%
  separate(elect, c('e1', 'e2'), sep = '_') %>%
  mutate(source = paste0(source, ' ', gsub('f','', e1)),
         target = paste0(target, ' ', e2)) %>%
  select(-e1, -e2) %>%
  knitr::kable()
```

<br>


> A **Sankey diagram** detailing transitions in presidential voting patterns (at the district level) from (1) 2008 to 2012 and (2) 2012 to 2016.  Clearly a bit jazzier as an [html widget proper](https://www.jtimm.net/odds/).


```{r message=FALSE, warning=FALSE}
library(plotly)
r<- '#9e5055'; b <- '#395f81'
#htmltools::save_html(viz, 'pres_sank.html')

viz <- plot_ly(
    type = "sankey",
    orientation = "h",
    node = list(
      label = c("McCain '08", "Obama '08", "Romney '12", 
                "Obama '12", "Trump '16", "Clinton '16"),
      color = c(r, b, r, b, r, b),
      pad = 15,
      thickness = 20,
      line = list(color = "black", width = 0.5)),
    
    link = list(
      source = c(0,0,1,1,3,3,2,2),
      target = c(3,2,3,2,5,4,5,4),
      value =  c(1,192,210,32,189,21,15,207))) %>% 
  
    layout(
      title = "Transitions in district-level presidential voting patterns: 2008-2016",
      font = list(size = 10))
viz
```

---


### 7 Summary


**Hopefully a nice round-up** of useful open source resources for investigating & visualizing some American political data.  I will update & develop as things get updated & develop.  Let us know!  

