---
output:
  md_document:
    variant: markdown_github
    toc: FALSE
---

##American political data & R: some open source resources & methods

A layman's guide to accessing, integrating, and exploring US political data from a collection of open government resources, including presidential election returns (2008-2016 by congressional district), lawmaker biographies & political ideologies, and congressional district demographics.

Data presented here have been collated from [The Daily Kos](), [CivilServiceUSA](), and the R packages [tidycensus]() & [Rvoteview]().


- [1 Lawmaker details](#1-Lawmaker-details)
- [2 Political Ideologies](#2-political-ideologies-and-congressional-composition)
- [3 Political geometries](#3-political-geometries)
- [4 Federal election results](#4-Federal-election-results)
- [5 Census data and congressional districts](#5-Census-data-and-congressional-districts)
- [6 Alternative geometries](#6-Alternative-geometries)
- [7 Summary](#7-Summary)


Hopefully a useful open source & transparent framework for investigating past & future election results and congresses using R.  All work presented here can be reproduced in its entirety. A developing resource.


```{r message=FALSE, warning=FALSE}
library(tidyverse)
```


---

### 1 Lawmaker details

> [CivilServiceUSA](https://github.com/CivilServiceUSA) provides a wonderful collection of details about each lawmaker in the 115th Congress, including age, race, religion, biographical details, and social media info.  A full roll call of information available for each lawmaker is available [here](https://github.com/CivilServiceUSA/us-house#data-set). Here, we consider some different perspectives on the composition of the 115th House utilizing these data.


```{r message=FALSE, warning=FALSE}
git <- 'https://raw.githubusercontent.com/CivilServiceUSA/'
csusa_house_dets <- jsonlite::fromJSON(url(paste0(git, 'us-senate/master/us-senate/data/us-senate.json')))
csusa_house_dets <- jsonlite::fromJSON(url(paste0(git, 'us-house/master/us-house/data/us-house.json')))
```



#### 1.1 Age & generational demographics of the 115th House

```{r fig.height=4.5, message=FALSE, warning=FALSE}
csusa_house_dets %>%
  mutate (years = 
            lubridate::year(as.Date(Sys.Date())) -
            lubridate::year(as.Date(date_of_birth))) %>%
  ggplot (aes(years)) +
  geom_histogram(bins=20, fill = 'steelblue', alpha = .85) +
  labs(title = '115th House composition by age',
       caption = 'Data source: CivilServiceUSA')
```


> [Pew Research](http://www.pewresearch.org/fact-tank/2018/04/11/millennials-largest-generation-us-labor-force/ft_15-05-11_millennialsdefined/) uses the following set of birth-year ranges to delineate generations.  

* Millenials: 1981-1997
* Generation X: 1965 -1980
* Baby Boomers: 1946-1964
* Silent: 1928-1945
* Greatest: < 1928


> For good measure, we provide a more detailed classification of Boomers --- lumping folks born post-WWII with those born in the ~Sixties is a problem for me. So, (a) Boomers-proper 1946-1954 & (b) [Generation Jones](https://en.wikipedia.org/wiki/Generation_Jones) 1955-1964. 


```{r fig.height=3}
gens115 <- csusa_house_dets %>%
  mutate (yob = as.numeric(gsub('-.*$', '', date_of_birth))) %>%
  mutate (gen = case_when (yob < 1998 & yob > 1980 ~ '4- Millenial',
                           yob < 1981 & yob > 1964 ~ '3- Gen X',
                           yob < 1965 & yob > 1954 ~ '2b - Gen Jones',
                           yob < 1955 & yob > 1945 ~ '2a - Boomer-proper',
                           yob < 1946 & yob > 1927 ~ '1 - Silent'))
gens115 %>%
  group_by(gen,party) %>%
  summarize(n=n()) %>%
  group_by(party) %>%
  mutate(rank = row_number())%>%
  ggplot(aes(x=reorder(gen, -rank), 
             y=n, 
             fill=gen)) + 
  geom_col(show.legend = FALSE, alpha = 0.85)+
  geom_text(aes(label = n), size = 3)+
  ggthemes::scale_fill_stata() +
  xlab(NULL) + ylab(NULL) +
  facet_wrap(~party) +
  coord_flip() +
  labs(title = '115th US House composition by generation',
       caption = 'Data source: CivilServiceUSA')
```




#### 1.2 Faith & the 115th House

```{r fig.height=5}
cols <- RColorBrewer::brewer.pal(4, 'Set1') 
cols <- colorRampPalette(cols)(31)

csusa_house_dets %>%
  group_by(religion) %>%
  summarize(n = n()) %>%
  na.omit() %>%
    ggplot(aes(area = n,
               fill = religion,
               label = religion,
               subgroup = religion)) +
      treemapify::geom_treemap(alpha=.85) +
      treemapify::geom_treemap_subgroup_border() +
      treemapify::geom_treemap_text(colour = "white", 
                        place = "topleft", 
                        reflow = T,
                        size = 11)+
      scale_fill_manual(values = cols) +
      theme(legend.position = "none",
            #plot.title = element_text(size=12),
            legend.title=element_blank()) +
      labs(title = '115th House composition by religion',
           caption = 'Data source: CivilServiceUSA')
```


---

### 2 Political ideologies and congressional composition

> The [VoteView](https://voteview.com/) project provides roll call-based political ideology scores for all lawmakers in the history of the US Congress.  Data can be used to investigate congressional composition by party affiliation over time, the aggregate political ideologies of both houses over time, and the ideologies of individual lawmakers. Plus a host of other roll call-based analyses.


```{r message=FALSE, warning=FALSE}
rvoteview_house_50 <- lapply(c(66:115), function (x)
                    Rvoteview::member_search (
                      chamber = 'House', 
                      congress = x)) %>% 
  bind_rows() %>%
  mutate(name = paste0 (gsub(', .*$', '', bioname), ' ', cqlabel))
```



#### 2.1 Congressional composition


```{r fig.height=5.5, message=FALSE, warning=FALSE}
rvoteview_house_50 %>%
  filter(party_name %in% c('Democratic Party', 'Republican Party')) %>%
  group_by(congress, party_name) %>%
  summarize(n = n()) %>%
  mutate(n = n/sum(n)) %>%
  ggplot(aes(x=congress, y=n, fill = party_name)) +
  geom_area(alpha = 0.85, color = 'gray') +
  ggthemes::scale_fill_stata()+
  geom_hline(yintercept = 0.5, color = 'white', linetype = 2) +
  annotate("text", x = 73, y = .9, label = "Democrats", 
           color = 'white',
           size = 4) +
  annotate("text", x = 73, y = .1, label = "Republicans", 
           color = 'white',
           size = 4) +
  theme(legend.position = "none")+
  labs(title = "House Composition over the last 50 congresses",
       caption = 'Data source: VoteView')
```


#### 2.2 Lawmaker political ideologies

> A brief description of methods. Interpreting ideology scores.

```{r}
extremes <- rvoteview_house_50 %>%
  filter(congress == 111) %>%
  select(name, party_name, nominate.dim1, nominate.dim2) %>%
  gather (key = dim, value = estimate,
          nominate.dim1:nominate.dim2) %>%
  group_by(party_name,dim) %>%
  filter(estimate == max((estimate)) | 
           estimate == min((estimate)))
```


> The 111th House took office in 2009; members were elected in a Presidential election year with Obama & McCain on the top of the ticket.  Below we plot members of the 111th House in two-dimensional political space.  Ideoligically extreme members are labelled.

```{r fig.height=5, message=FALSE, warning=FALSE}
rvoteview_house_50 %>%
  filter (congress == 111) %>%
  ggplot(aes(x=nominate.dim1, 
             y=nominate.dim2, 
             label = name
             )) +
          annotate("path",
               x=cos(seq(0,2*pi,length.out=300)),
               y=sin(seq(0,2*pi,length.out=300)),
               color='gray',
               size = .25) +
  geom_point(aes(color = as.factor(party_code)), 
             size= 2, 
             shape= 17) +
  ggrepel::geom_text_repel(
    data  = filter(rvoteview_house_50, 
                   congress==111,
                   name %in% extremes$name),
    nudge_y =  -0.025,
    direction = "y",
    hjust = 0, size = 2.5) +
  ggthemes::scale_color_stata() +
  theme(legend.position = 'none') +
  labs(title="DW-Nominate ideology scores for the 111th US House",
       caption = 'Data source: VoteView')
```




#### 2.3 Political ideologies historically: a party-based overview

> A slightly modified version of [this](https://voteview.com/parties/all) VoteView visualization illustrating the growing ideological divide between major political parties in the US. The gray line reflects overall median ideology scores in the House, and (albeit indirectly) party control of the House historically.

```{r fig.height=5.5}
rvoteview_house_50 %>%
  filter(!is.na(nominate.dim1) & 
           party_name %in% c('Democratic Party','Republican Party')) %>%
  group_by(congress) %>%
  mutate(xmed = median(nominate.dim1)) %>%
  group_by(congress, party_name, xmed) %>%
  summarize(med = median(nominate.dim1)) %>%
  ungroup() %>%
  mutate(year = 1916 + 2*rep(c(1:50), each = 2)) %>%
  ggplot() +
  geom_line(aes(x = year, y= med, color = party_name), size = 1) +
  ggthemes::scale_color_stata()+
  geom_line(aes(x = year, y= xmed), color = 'darkgray',  size = 1.25) +
  theme(legend.position = 'bottom') +
  labs(title="Median ideologies for major parties: Houses 66 to 115",
       caption = 'Data source: VoteView') 
```




#### 2.4 NOKKEN & POOLE scores

> An alternative set of ideology scores are available as a csv from VoteView's website.  DW-Nominate scores are constant over a lawmakers entire political career.  Nokken & Poole scores, in contrast, are congress-specific scores.


```{r message=FALSE, warning=FALSE, eval=FALSE}
voteview_nokken_poole <- read.csv(url("https://voteview.com/static/data/out/members/HSall_members.csv"),
  stringsAsFactors = FALSE) 
```


---


### 3 Political geometries

> The R package `tigris` provides access to all US Census-based geometries, including US congressional districts and state legislative districts.

```{r message=FALSE, warning=FALSE}
nonx <- c('78', '69', '66', '72', '60', '15', '02')

library(tigris); options(tigris_use_cache = TRUE, tigris_class = "sf")
us_house_districts <- tigris::congressional_districts(cb = TRUE) %>%
  select(GEOID,STATEFP, CD115FP) %>%
  
  left_join(tigris::states(cb = TRUE) %>% 
              data.frame() %>%
              select(STATEFP, STUSPS)) 

laea <- sf::st_crs("+proj=laea +lat_0=30 +lon_0=-95") # Lambert equal area
us_house_districts <- sf::st_transform(us_house_districts, laea)
```

  
---

### 4 Federal election results

> The Daily Kos makes available [a host of data sets](https://www.dailykos.com/stories/2018/2/21/1742660/-The-ultimate-Daily-Kos-Elections-guide-to-all-of-our-data-sets), including federal election returns for presidential & congressional races.  2016 returns are the most recent data sets available. Here we focus on presidential results for the last three elections.  Seemingly one of the more comprehensive collections of open source federal election returns.



#### 4.1 Restructuring election data

```{r message=FALSE, warning=FALSE}
url <- 'https://docs.google.com/spreadsheets/d/1oRl7vxEJUUDWJCyrjo62cELJD2ONIVl-D9TSUKiK9jk/edit#gid=1178631925'
house <- gsheet::gsheet2tbl(url) 
```



> A simple cleaning procedure that should (more or less) scale to other data sources at the Daily Kos. 

```{r message=FALSE, warning=FALSE}
fix <- as.data.frame(cbind(colnames(house), as.character(house[1,])), 
  string_as_factor = FALSE) %>%
  mutate(V1 = gsub('^X', NA, V1)) %>%
  fill(V1) %>%
  mutate(nw_cols = ifelse(is.na(V2), V1, paste0(V1, '_', V2)),
         nw_cols = gsub(' ', '_', nw_cols))

colnames(house) <- fix$nw_cols
house <- house %>% slice(3:nrow(.))
keeps <- house[,!grepl('Pronun|ACS|Census|Survey', colnames(house))]
```



> Our new data structure, filtered to Presidential returns.

```{r message=FALSE, warning=FALSE}
dailykos_pres_elections <- keeps [,c('District', 'Code', grep('President_[A-z]', colnames(house), value=T))] %>%
  gather (key = election, value = percent, `2016_President_Clinton`:`2008_President_McCain`) %>%
  mutate(election = gsub('President_', '', election),
         percent = as.numeric(percent)) %>%
  separate(Code, c('STUSPS', 'CD115FP')) %>%
  separate(election, c('year', 'candidate'))%>%
  mutate(CD115FP = ifelse(CD115FP == 'AL', '00', CD115FP)) %>%
  left_join(data.frame(us_house_districts) %>% select (-geometry))
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
dailykos_pres_elections %>%
  select(District, GEOID, year, candidate, percent) %>%
  head() %>%
  knitr::kable()
```




#### 4.2 Presidential Election results - 2016

```{r fig.height=6, message=FALSE, warning=FALSE}
us_house_districts %>%
  filter(!gsub('..$' ,'', GEOID) %in% nonx) %>%
  left_join(dailykos_pres_elections %>% 
              filter(year == '2016') %>%
              spread(candidate, percent) %>%
              mutate(Margin = Trump-Clinton)) %>%
  ggplot() + 
  geom_sf(aes(fill = Margin)) +
  scale_fill_distiller(palette = "RdBu",direction=-1)+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        legend.position = 'bottom') +
  labs(title = "Trump vote margins by congressional district",
       caption = 'Data source: Daily Kos')
```



#### 4.3 Rural & urban voting

> Using the area of congressional districts (in log square meters) as a proxy for the often cited urban-rural divide in American votership, the plot below illustrates the relationship between Trump margins and ~degree of urbanicity.  The pattern below helps account for the very red map above despite Clinton's success in the popular vote.

```{r fig.height=5, message=FALSE, warning=FALSE}
us_house_districts %>%
  left_join(dailykos_pres_elections %>% 
              filter(year == '2016') %>%
              spread(candidate, percent) %>%
              mutate(Margin = Trump-Clinton)) %>%
  mutate(area = as.numeric(gsub(' m^2]', '', sf::st_area(.))),
         party = ifelse(Margin > 0, 'Trump', 'Clinton')) %>%
  ggplot(aes(y=Margin, x=log(area), color = party)) +
  geom_point(alpha = .65) +
  ggthemes::scale_color_stata()+
  geom_smooth(method="lm", se=T, color = 'black', linetype = 3)+
  geom_hline(yintercept = 0, color = 'darkgrey') +
  theme(legend.position = "none")+
  labs(title = "Trump vote margins vs. log(area) of congressional district",
       caption = 'Data source: Daily Kos')
```


---

### 5 Census data and congressional districts

> The US Census/American Community Survey (ACS) make counts/estimates available by congressional district.  The R package `tidycensus` provides access to census APIs.  Here we gather educational attainment data.  

```{r message=FALSE, warning=FALSE, include=FALSE}
var_list <- tidycensus::load_variables(2017, "acs5")
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
tidycensus::census_api_key("b508704c99f3ae9bc5b5e7c41e3dd77e59d52722")
```



#### 5.1 Educational attainment by race/ethnicity

```{r message=FALSE, warning=FALSE}
code <- c('A', 'B', 'C', 'D', 'E',
          'F', 'G', 'H', 'I')
          
          
race <- c('WHITE ALONE', 'BLACK OR AFRICAN AMERICAN ALONE',
          'AMERICAN INDIAN OR ALASKAN NATIVE ALONE',
          'ASIAN ALONE', 
          'NATIVE HAWAIIAN AND OTHER PACIFIC ISLANDER ALONE', 
          'SOME OTHER RACE ALONE', 'TWO OR MORE RACES',
          'WHITE ALONE, NOT HISPANIC OR LATINO',
          'HISPANC OR LATINO')

race_table <- as.data.frame(cbind(code,race),
                            stringsAsFactors=FALSE)
```



> Table C15002: *Sex by educational attainment for the population 25 years and over*.  Data disaggregated by race & ethncity can be accessed via suffixes A-I. The table above crosses suffix to subgroup classification.

```{r message=FALSE, warning=FALSE}
search_vars <- var_list[grepl('C1500', var_list$name),]

tidycens_data <- tidycensus::get_acs(geography = 'congressional district',
                            variables = search_vars$name,
                            summary_var = 'B15002_001',
                            year = 2017,
                            survey = 'acs5') %>%
  left_join(search_vars %>% rename(variable = name)) %>%
  filter(!grepl('Total$|Female$|Male$', label)) %>%
  
  mutate(gender = ifelse(grepl('Male', label), 'Male', 'Female'),
         label = gsub('^Estimate.*!!', '', label),
         code = gsub('(C[0-9]+)([A-Z])(_[0-9]+.$)', 
                     '\\2', 
                     variable)) %>%
  left_join (race_table) %>%
  select(GEOID, label, gender, race, estimate:summary_moe)
```



#### 5.2 The White working class

> In the Trump era, educational divides in voting behavior have received a great deal of attention.  In particular, pundits & news outlets have focused on a segment of the population they have dubbed the "White working class."  

> **White working class** formalized in US Census terms: Population 25 years & older who identify as White & non-Hispanic without a Bachelor's degree.


```{r fig.height=6, message=FALSE, warning=FALSE}
us_house_districts %>% 
  filter(!gsub('..$' ,'', GEOID) %in% nonx) %>%
  left_join(tidycens_data %>% 
              filter(label != 'Bachelor\'s degree or higher' &
                     #gender == 'Male' & 
                     race == 'WHITE ALONE, NOT HISPANIC OR LATINO')%>%
              group_by(GEOID) %>%
              summarize(estimate = sum(estimate), 
                        summary_est = mean(summary_est))) %>%
  mutate(per = estimate / summary_est) %>%
  ggplot() + 
  geom_sf(aes(fill = per)) + 
  
  scale_fill_distiller(palette = "BrBG", direction=1)+
  
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        legend.position = 'bottom') +
  labs(title = "White working class (%) by congressional district",
       caption = 'Data source: American Community Survey, 5-Year estimates, 2013-17, Table C15002')
```



#### 5.3 Educational attainment profiles for congressional districts

> Often ... these comparisons ... not formal census categories, and require some aggregation.

```{r message=FALSE, warning=FALSE}
tree <- tidycens_data %>%
  left_join(data.frame(us_house_districts) %>% select(GEOID, STUSPS, CD115FP)) %>%
  mutate (race = gsub(', | ', '_', race)) %>%
  select(-moe:-summary_moe) %>%
  spread(race, estimate) %>%
  mutate(WHITE_ALONE_HISPANIC = WHITE_ALONE - WHITE_ALONE_NOT_HISPANIC_OR_LATINO) %>%
  gather(key =race, value = estimate, AMERICAN_INDIAN_OR_ALASKAN_NATIVE_ALONE:WHITE_ALONE_HISPANIC) %>%
  filter(race != 'HISPANIC OR LATINO') %>%
  mutate(race_cat = ifelse(race == 'WHITE_ALONE_NOT_HISPANIC_OR_LATINO', 'White', 'Non-White'),
    ed_cat = ifelse(label == 'Bachelor\'s degree or higher', 'College', 'Non-College'))%>%
  group_by(GEOID, STUSPS, CD115FP, race_cat, ed_cat) %>%
  summarize(estimate = sum(estimate)) %>%
  group_by(GEOID) %>%
  mutate(per = estimate/sum(estimate)) %>%
  ungroup()
```


> A random sample of congressional districts:

```{r fig.height=6, message=FALSE, warning=FALSE}
samp_n <- sample(unique(tree$GEOID), 12)

tree %>%
  filter(GEOID %in% samp_n) %>%
    ggplot(aes(area = per,
               fill = paste0(race_cat, ' ', ed_cat),
               label = paste0(race_cat, ' ', ed_cat),
               subgroup = paste0(race_cat, ' ', ed_cat)))+
      treemapify::geom_treemap(alpha=.85)+
      treemapify::geom_treemap_subgroup_border() +

      treemapify::geom_treemap_text(colour = "white", 
                        place = "topleft", 
                        reflow = T,
                        size = 9)+
      #ggthemes::scale_fill_stata()+ 
      scale_fill_brewer(palette = 'Paired') +
      facet_wrap(~paste0(STUSPS, '-', CD115FP)) +
      theme(legend.position = "bottom",
            legend.title=element_blank()) + 
      labs(title = "Educational attainment by race for population over 25",
           subtitle = 'A random sample of congressional districts',
           caption = 'Source: American Community Survey, 5-Year estimates, 2013-17, Table C15002')
```



#### 5.4 Trump support by educational attainment

> Thoughts.

```{r fig.height=6, message=FALSE, warning=FALSE}
by_pres <- dailykos_pres_elections %>%
  filter(candidate %in% c('McCain', 'Romney', 'Trump')) %>% 
  left_join(tree %>% filter(race_cat == 'White' & 
                              ed_cat == 'Non-College')) %>%
  rename(Per_White_Working = per, Per_Rep_Margin = percent)
  
ggplot(data = by_pres, aes(x=Per_White_Working, 
                           y=Per_Rep_Margin, 
                           fill = candidate, 
                           color = candidate, 
                           linetype = candidate)) +
  geom_point(alpha = .75, size = 1) +
  ggthemes::scale_fill_stata()+
  ggthemes::scale_color_stata()+
  geom_smooth(method="lm", se=T, color = 'black', size = .5)+
  theme(legend.position = "bottom")+
  labs(title = "% White working class vs. % Republican support",
       subtitle = 'By Presidential election',
       caption = 'Data source: Daily Kos & American Community Survey')
```


> So, this relationship has grown stronger in the Trump era.

```{r}
by_pres %>% 
  group_by(candidate) %>%
  summarize(cor(Per_White_Working, Per_Rep_Margin)) %>%
  knitr::kable()
```


---

### 6 Alternative political geometries

> The Daily Kos makes available of set of shapefiles meant to represent congressional districts and states as ....  The Daily Vos makes these shapefiles availble via Google Drive.  Links are provided below.


```{r}
base <- 'https://drive.google.com/uc?authuser=0&id='
#Hex map
dailyvos_hex_cd <- paste0(base, '1E_P0r1Uv438fZsvKsvidIR02Nb5Ju9zf&export=download/HexCDv12.zip')
dailyvos_hex_st <- paste0(base, '0B2X3Bx1aCHsJVWxYZGtxMGhrMEE&export=download/HexSTv11.zip')
#Tile map
dailyvos_tile_outer <- paste0(base, '0B2X3Bx1aCHsJdGF4ZWRTQmVyV2s&export=download/TileOutv10.zip')
dailyvos_tile_inner <- paste0(base, '0B2X3Bx1aCHsJR1c0SzNyWlAtZjA&export=download/TileInv10.zip')
```



#### 6.1 A simple function for shapefile extraction

```{r message=FALSE, warning=FALSE}
get_url_shape <- function (url) {
  temp <- tempdir()
  zip_name <- paste0(temp, '\\', basename(url))
  download.file(url, zip_name, 
                quiet = TRUE)
  unzip(zip_name, exdir = temp)
  x <- sf::st_read(dsn = gsub('\\.zip', '', zip_name), 
                   layer = gsub('\\.zip','', basename(url)),
                   quiet = TRUE) 
  unlink(temp) 
  x}
```


#### 6.2 Tile map of US states: Senate composition historically

> 

```{r}
dailykos_tile <- lapply (c(dailyvos_tile_inner,
                           dailyvos_tile_outer),
                         get_url_shape)
names(dailykos_tile) <- c('inner', 'outer')
```


```{r message=FALSE, warning=FALSE, include=FALSE}
rvoteview_senate_50 <- lapply(c(66:115), function (x)
                    Rvoteview::member_search (
                      chamber = 'Senate', 
                      congress = x)) %>% 
  bind_rows() 
```


> Here we 

```{r}
sens <- rvoteview_senate_50 %>%
  mutate(party_name = factor(party_name, levels =c('Democratic Party', 
                                         'Republican Party', 
                                         'Independent'))) %>%
  filter(congress %in% c(85, 91, 97, 103, 109, 115)) %>%
  arrange (state_abbrev, party_name) %>%
  group_by(congress, state_abbrev) %>%
  mutate(layer = row_number())%>%
  rename(State = state_abbrev) %>%
  select(congress, State, party_name, layer)
```



> The tile map below demonstrates the evolution of US Senate composition by state and party affiliation over the last thirty congresses.

```{r fig.height=6, message=FALSE, warning=FALSE}
dailykos_tile$outer %>% 
  left_join(sens %>% filter (layer == 2)) %>%
  ggplot() + 
  geom_sf(aes(fill = party_name),
          color = 'black', 
          alpha = .85) + 
  geom_sf(data = dailykos_tile$inner %>%
            left_join(sens %>% filter (layer == 1)), 
          aes(fill = party_name)) +
  ggsflabel::geom_sf_text(data = dailykos_tile$inner,
                          aes(label = State), 
                          size = 1.75,
                          color = 'white') +
  ggthemes::scale_fill_stata()+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        legend.title=element_blank(),
        legend.position = 'bottom') +
  facet_wrap(~congress) +
  labs(title = "US Senate Composition by Congress, State & Party: Snapshots",
       caption = 'Data sources: Daily Kos & VoteView')
```



#### 6.3 Hexmap of Congressional districs: Presidential lineages

```{r message=FALSE, warning=FALSE}
dailykos_shapes <- lapply (c(dailyvos_hex_cd, dailyvos_hex_st), 
                           get_url_shape)
names(dailykos_shapes) <- c('cds', 'states')
#State hex shapefile is slightly broken.
dailykos_shapes$states <- lwgeom::st_make_valid(dailykos_shapes$states)
```


>   

```{r}
dailykos_pres_flips <- dailykos_pres_elections %>%
  group_by(District, year) %>%
  filter(percent == max(percent))%>%
  
  mutate(dups = n()) %>%
  filter(dups != 2) %>% #Kill ties --> n = 3
  
  select(-percent, -dups) %>% #
  arrange(District) %>%
  spread(year, candidate) 

#Some hand edits to address percentage "ties" 
dailykos_pres_flips$`2012`[dailykos_pres_flips$District == 'Florida 7th'] <- 'Obama' 
dailykos_pres_flips$`2008`[dailykos_pres_flips$District == 'Ohio 10th'] <- 'Obama' 
dailykos_pres_flips$`2008`[dailykos_pres_flips$District == 'New York 22nd'] <- 'McCain'
```



```{r}
dailykos_pres_flips <- dailykos_pres_flips %>%
  mutate(flips = paste0(`2008`, '-',`2012`, '-', `2016`)) %>%
  group_by(flips) %>%
  mutate(sum = n()) %>%
  ungroup()
```


Summary
```{r fig.height=3.5, message=FALSE, warning=FALSE}
dailykos_pres_flips %>%
  group_by(flips) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  ggplot(aes(x=reorder(flips, n), 
             y=n, 
             fill=reorder(flips, -n))) + 
  geom_col(show.legend = FALSE, alpha = 0.85)+
  geom_text(aes(label = n), size = 3)+
  ggthemes::scale_fill_stata() +
  xlab(NULL) + ylab(NULL) +
  coord_flip() +
  labs(title = 'Presidential winner lineages - 2008-16',
       caption = 'Data source: Daily Kos')
```


Note that this has been reproduced.

```{r fig.height=6, message=FALSE, warning=FALSE}
dailykos_shapes$cds %>%
  inner_join(dailykos_pres_flips)%>%
  ggplot() + 
  geom_sf(aes(fill = reorder(flips, -sum)),
          color = 'gray', 
          alpha = .85) + 
  geom_sf(data=dailykos_shapes$states, 
          fill = NA, 
          show.legend = F, 
          color="black", 
          lwd=.7) +
  ggsflabel::geom_sf_text(data = dailykos_shapes$states,
                          aes(label = STATE), 
                          size = 2.75,
                          color='white',
                          face='bold') +
  ggthemes::scale_fill_stata()+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        legend.title=element_blank(),
        legend.position = 'bottom') +
  labs(title = "Presidential election results by district - 2008, 2012 & 2016",
       caption = 'Data source: Daily Kos')
```




#### 6.4 Another perspective

```{r}
dailykos_pres_flips %>%
  mutate(f08_12 = paste0(`2008`,'_', `2012`),
         f12_16 = paste0(`2012`,'_', `2016`))%>%
  select(District, f08_12, f12_16) %>%
  gather(elect, flip, -District) %>%
  group_by(elect, flip) %>%
  summarize(value=n()) %>%
  separate(flip, c('source', 'target'), sep = '_') %>%
  separate(elect, c('e1', 'e2'), sep = '_') %>%
  mutate(source = paste0(source, ' ', gsub('f','', e1)),
         target = paste0(target, ' ', e2)) %>%
  select(-e1, -e2) %>%
  knitr::kable()
```



A sankey diagram.  Clearly a bit jazzier as an html widget proper.

```{r message=FALSE, warning=FALSE}
library(plotly)
r<- '#9e5055'; b <- '#395f81'
plot_ly(
    type = "sankey",
    orientation = "h",
    node = list(
      label = c("McCain '08", "Obama '08", "Romney '12", 
                "Obama '12", "Trump '16", "Clinton '16"),
      color = c(r, b, r, b, r, b),
      pad = 15,
      thickness = 20,
      line = list(color = "black", width = 0.5)),
    
    link = list(
      source = c(0,0,1,1,3,3,2,2),
      target = c(3,2,3,2,5,4,5,4),
      value =  c(1,192,210,32,189,21,15,207))) %>% 
  
    layout(
      title = "Presidential support by Congressional District count",
      font = list(size = 10))
```


---


### 7 Summary


Indeed, a work in progress.  But hopefully a nice round-up of useful open source resources for investigating & visualizing federal election results.  

